```{r include = FALSE, echo = FALSE, output = FALSE, warning = FALSE}
library(skimr)
library(kableExtra)
library(tidyverse)
library(tidycensus)
```

```{r echo = FALSE, output = TRUE}

coal <-
  read_rds('data/raw_coal_data.rds')

# skim(coal)

fips <-
  tidycensus::fips_codes %>%
  select(
    state,
    state_code,
    state_name) %>%
  summarise(.by = state)
```

So we've got this huge dataset of 220k coal mining incidents that were reported.

I plan to do a deep dive into everything and figure out what I can, as per the showcase plan memo (within reason/practicality; Shiny is proving to be more annoying than I could possibly have anticipated).

Let's just look at the "top 5" in table form for now in a few categories. We'll look at companies, states, injured body parts, and injury sources.

```{r include = FALSE, echo = FALSE, output = FALSE, warning = FALSE}
top_fives <- 
  list(
  'OPERATOR_NAME',
  'FIPS_STATE_CD',
  'INJ_BODY_PART',
  'INJURY_SOURCE') %>%
  map(
    ~ coal %>%
      pull(.x) %>%
      table() %>%
      as.data.frame() %>%
      arrange(desc(Freq)) %>%
      head(5) %>%
      set_names(
        .x,
        'Freq') %>%
      kbl() %>%
      kable_paper(
        "hover",
        full_width = FALSE))
```

```{r echo = FALSE}
# I hate how WET this code is, but I'm on a deadline.
# Don't judge me.
top_fives[[1]]
top_fives[[2]]
top_fives[[3]]
top_fives[[4]]
```

I plan on looking at MANY more things over the coming days, but that's what I could do on a quick-turnaround. Stay tuned!
